---
title: "Association of Smoking with Cost in a Simulated Dataset for HRP 203"
author:
  - name: Vikram Fielding-Singh, MD, JD
    affiliations:
      - name: Department of Anesthesiology, Perioperative and Pain Medicine, Stanford University School of Medicine
format: 
  docx:
    toc: false
    number-sections: false
    highlight-style: github
editor: visual
bibliography: HRP203.bib
csl: american-medical-association.csl
---

## Background

Smoking is associated with numerous cardiovascular risk factors.[@virdis_cigarette_2010; @prochaska_smoking_2015] It may also be associated increased health care costs, although some have suggested that smoking cessation will be associated with increased health care costs.[@barendregt_health_1997] In order to practice using reproducible methods, we conducted a study to evaluate whether smoking is associated with increased health care costs in a simulated dataset.

```{r}
#| label: package-load
#| echo: false
#| warning: false

#Load packages
pacman::p_load(pacman, tidyverse, rmarkdown, readxl, table1, ggplot2, geepack)

#Set home directory and load here after tidyverse to avoid conflicts
here::i_am("README.md")
p_load(here)
```

```{r}
#| label: data-load
#| echo: false
#| warning: false

#Load data
cohort <- read.csv(here("raw-data","cohort.csv"))
```

## Methods

```{r}
#| label: methods
#| echo: false
#| warning: false
```

We analyzed a cohort of simulated observations with the following variables: `smoke`, `female`, `age`, `cardiac`, and `cost`.

We first report descriptive statistics for the cohort and analyze the unadjusted relationship between smoking and cost. To account for possible confounders, we also fit a linear regression model:

$cost_i = \beta_0 + \beta_1 smoke_i + \beta_2 female_i + \beta_3 age_i + \beta_4 cardiac_i + \epsilon_i$

for $i = 1,...,n$. Analysis was conducted on June 1, 2025.

## Results

```{r}
#| label: results_descriptive
#| echo: false
#| warning: false

# Counts to use in the prose
female_counts <- cohort %>% count(female)
female_pct <- female_counts *100 / nrow(cohort)

#Cost by groups
cost_by_smoker_mean <- cohort %>% group_by(smoke) %>% summarize(Mean = mean(cost))
cost_by_gender_mean <- cohort %>% group_by(female) %>% summarize(Mean = mean(cost))
cost_by_gender_sd <- cohort %>% group_by(female) %>% summarize(SD = sd(cost))
```

There were `{r} format(nrow(cohort), big.mark = ",")` simulated observations, of whom `{r} format(female_counts[2,2], big.mark = ",")` / `{r} format(nrow(cohort), big.mark = ",")` (`{r} format(round(female_pct[2,2], digits = 1), big.mark = ",")`%) were female [(Table 1)](#table_1). The mean age was `{r} format(round(mean(cohort$age), digits = 1), big.mark = ",")` (standard deviation `{r} format(round(sd(cohort$age), digits = 1), big.mark = ",")`) years. The average cost in the cohort was \$`{r} round(mean(cohort$cost), digits = 0)` (standard deviation \$`{r} round(sd(cohort$cost), digits = 0)`), with an average cost of \$`{r} round(cost_by_smoker_mean[2,2], digits = 0)` among smokers and \$`{r} round(cost_by_smoker_mean[1,2])` among nonsmokers [(Figure 1)](#figure_1). Average costs were \$`{r} round(cost_by_gender_mean[1,2])` (standard deviation \$`{r} round(cost_by_gender_sd[1,2])`) among males and \$`{r} round(cost_by_gender_mean[2,2])` (standard deviation \$`{r} round(cost_by_gender_sd[2,2])`) among females [(Figure 2)](#figure_2). Finally, increasing age was associated with increased cost [(Figure 3)](#figure_3).

```{r}
#| label: results_regression
#| echo: false
#| warning: false

#Regression
adj_model <- lm(cost ~ age + female + smoke + cardiac, data=cohort)
coefs <- coef(adj_model)
cis <- confint(adj_model)
```

In adjusted analysis, smoking was associated with \$`{r} round(coefs[4])` (95% confidence interval \$`{r} round(cis[4,1])` to \$`{r} round(cis[4,2])`) of additional cost. 

## Discussion
We report a brief analysis examining the relationship between smoking and cost in a simulated dataset. In both unadjusted and adjusted analyses, smoking was associated with higher costs. We also demonstrate several features of reproducible reports, including in-line citations using citation style language to specify the format, relative file paths, embedded code for analyses with linked tables and figures, and inline math notation.

## Generative AI Statement
No generative AI technology was used to complete any portion of this assignment.

## References
::: {#refs}
:::

{{< pagebreak >}}

## Tables

###### Table 1. Cohort Characteristics by Smoking Status {#table_1}
```{r}
#| label: table1
#| echo: false
#| warning: false

#Some formatting
cohort$smoke <- 
  factor(cohort$smoke, levels=c(0,1),
         labels=c("Nonsmoker", 
                  "Smoker"))
cohort$female <- 
  factor(cohort$female, levels=c(0,1),
         labels=c("Male", 
                  "Female"))

cohort$cardiac <- 
  factor(cohort$cardiac, levels=c(0,1),
         labels=c("No", 
                  "Yes"))
label(cohort$age)       <- "Age"
label(cohort$female)    <- "Gender"
label(cohort$cost)      <- "Cost"
label(cohort$cardiac)   <- "Cardiac"
label(cohort$smoke)     <- "Smoker"

#Use table1 package because it's pretty straightforward
table1(~age + female + cardiac + cost | smoke,
       data=cohort, overall=c(left="Total"))

```


## Figures

###### Figure 1. Cost by Smoking Status {#figure_1}
```{r}
#| label: figure1
#| echo: false
#| warning: false

#Cost by smoking status
ggplot(cohort, aes(x = smoke, y = cost, fill = smoke)) + 
  geom_boxplot(alpha = 0.5, show.legend = FALSE) + #Make colors slightly transparent and suppress legend
  labs(
    x = "Smoking Status",
    y = "Cost (USD)",
    title = ""
  ) +
  theme_minimal() +  # White background
  scale_fill_manual(values = c("blue", "darkgreen")) +
  theme(
    plot.title = element_text(hjust = 0.5),  # Center the title
  )

```

###### Figure 2. Cost by Gender {#figure_2}
```{r}
#| label: figure2
#| echo: false
#| warning: false

#Cost by gender
ggplot(cohort, aes(x = female, y = cost, fill = female)) + 
  geom_boxplot(alpha = 0.5, show.legend = FALSE) + #Make colors slightly transparent and suppress legend
  labs(
    x = "Female",
    y = "Cost (USD)",
    title = ""
  ) +
  theme_minimal() +  # White background
  scale_fill_manual(values = c("blue", "darkgreen")) +
  theme(
    plot.title = element_text(hjust = 0.5),  # Center the title
  )

```

###### Figure 3. Age versus Cost {#figure_3}
```{r}
#| label: figure3
#| echo: false
#| warning: false
#| 
#Age and cost scatter
ggplot(cohort, aes(x = age, y = cost)) + 
  geom_point(color = "blue", alpha = 0.7) +
  labs(
    x = "Age (years)",
    y = "Cost (USD)",
    title = ""
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),  # Center the title
  )

```

