---
title: "HRP203 Assignment 2"
author: "Vikram Fielding-Singh"
format: html
editor: visual
---

# Assignment 2

This file executes a short analysis for HRP 203, Assignment 2. No generative AI technology was used to complete any portion of this work.

The raw data is a cohort of 5000 simulated observations with the following variables: `smoke`, `female`, `age`, `cardiac`, and `cost`.

## Load Packages

```{r}
#| label: package-load
#| echo: false
#| warning: false

#Load packages
pacman::p_load(pacman, tidyverse, rmarkdown, readxl, table1, ggplot2, geepack)
```

## Load Dataset

```{r}
#| label: data-load

#Load data
cohort <- read.csv("D:/VFS Classes/HRP 203/ReproMethodsAssignment2/raw-data/cohort.csv")
```

## Quick Look

Have a quick look at the data.

```{r}
#| label: look

#Quick look
head(cohort)

# Continuous variables - look at distribution and summary statistics, check for missing
hist(cohort$age, breaks = 100)
summary(cohort$age)
hist(cohort$cost)
summary(cohort$cost)
table(is.na(cohort$age))
table(is.na(cohort$cost))

# Categorical variables
table(cohort$female, useNA = "ifany")
table(cohort$smoke, useNA = "ifany")
table(cohort$cardiac, useNA = "ifany")


```

## Table 1

Generate a Table 1 - will treat smoking as the primary exposure and cost as the primary outcome.

```{r}
#| label: table1

#Some formatting
cohort$smoke <- 
  factor(cohort$smoke, levels=c(0,1),
         labels=c("Nonsmoker", 
                  "Smoker"))
cohort$female <- 
  factor(cohort$female, levels=c(0,1),
         labels=c("Male", 
                  "Female"))

cohort$cardiac <- 
  factor(cohort$cardiac, levels=c(0,1),
         labels=c("No", 
                  "Yes"))
label(cohort$age)       <- "Age"
label(cohort$female)    <- "Gender"
label(cohort$cost)      <- "Cost"
label(cohort$cardiac)   <- "Cardiac"
label(cohort$smoke)     <- "Smoker"

#Use table1 package because it's pretty straightforward
table1(~age + female + cardiac + cost | smoke,
       data=cohort, overall=c(left="Total"))

```

## Figures

Generate some figures showing the relationships between variables of interest.

```{r}
#| label: figure1

#Cost by smoking status
ggplot(cohort, aes(x = smoke, y = cost, fill = smoke)) + 
  geom_boxplot(alpha = 0.5, show.legend = FALSE) + #Make colors slightly transparent and suppress legend
  labs(
    x = "Smoking Status",
    y = "Cost (USD)",
    title = "Cost by Smoking Status"
  ) +
  theme_minimal() +  # White background
  scale_fill_manual(values = c("blue", "darkgreen")) +
  theme(
    plot.title = element_text(hjust = 0.5),  # Center the title
  )

#Cost by gender
ggplot(cohort, aes(x = female, y = cost, fill = female)) + 
  geom_boxplot(alpha = 0.5, show.legend = FALSE) + #Make colors slightly transparent and suppress legend
  labs(
    x = "Female",
    y = "Cost (USD)",
    title = "Cost by Gender"
  ) +
  theme_minimal() +  # White background
  scale_fill_manual(values = c("blue", "darkgreen")) +
  theme(
    plot.title = element_text(hjust = 0.5),  # Center the title
  )

#Age and cost scatter
ggplot(cohort, aes(x = age, y = cost)) + 
  geom_point(color = "blue", alpha = 0.7) +
  labs(
    x = "Age (years)",
    y = "Cost (USD)",
    title = "Age versus Cost",
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),  # Center the title
  )


```

## Regression-Based Approach

Run a quick linear regression to test the adjusted relationship between `smoke` and `cost`.

```{r}
#| label: regression

#Regression
adj_model <- lm(cost ~ age + female + smoke + cardiac, data=cohort)
coef(adj_model)
confint(adj_model)

```
